% \subsection{Pre-computation of mismatch values for hamming distance computation (IHD)}
\subsection{Improved Hamming distance computation by pre-computation of mismatch values (IHD)}
The EMS-GT2 uses the hamming distance computation heavily during the Test phase. As discussed earlier for the original EMS-GT, the hamming distnace of two binary represented $l$-mers, can be efficiently computed usign the boolean operator XOR. In EMS-GT2, instead of repeatedly counting this non-zero pairs of bits every time we compute the hamming distance, we use a pre-computed lookup table to help reduce computational time. Figure \ref{fig:pre-computed-hd} shows a snapshot of this table of pre-computed number of non-zero pairs of bits.

\input{contents/00_figure/pre-computed-hd}

A naive pre-computation of these nonzero pair counts for all possible $l$-mers (each represented by 2$l$ bit values) will introduce an unacceptable overhead computation time when $l$ is sufficiently large, i.e., when $l \geq 10$ (based on actual runs on our current machine configurations). A more efficient approach is to pre-compute only up to $l$-mers of length $l' < l$, which required $b = 2l'$ number of bits. Then we determine the hamming distance by looking up to the nonzero counts in the XOR results, b number of bits at a time, as described in Algorithm 1. To illustrate the new process, Figure \ref{fig:usage-improved-hd} shows a demonstration of the improved hamming distance computation.

\input{contents/00_pseudocodes/improved-xor-computation} 

\input{contents/00_figure/usage-improved-hd}

In our experimentation the maximum required bits to represent $l$-mers is 34 bits (for (17, 6)-instance). Given this, we pre-compute up to 18 bits ($l' = 9$) values only and use the lookup table twice for the computation of the actual hammign distance between any given pair of $l$-mers.







