
\begin{abstract}\boldmath 

Finding DNA motifs is a widely studied area in the field of Computational Biology. Motifs signify different information that are useful for biologists like discovering transcription factor-binding sites and helps them identify potential therapeutic drug target. There are several variations of the motif finding problem, and one of these is called the $(l, d)$-motif search or Planted Motif Search problem (PMS). In this paper, we propose the EMS-GT2 algorithm, an extension of the Exact Motif Search - Generate and Test (EMS-GT) which is an exact enumerative algorithm for PMS. In EMS-GT2, we incorporated a new speedup technique that is based on an important property that we have discovered, which we prove in this paper, and which has enabled a more efficient block-processing of candidate motifs. We also improved the Hamming distance computation in EMS-GT2 by pre-computing mismatch values. We also explored a way to improve the $d$-neighborhood generation by focusing on non empty blocks in the candidate motifs array. Our C++ implementation of EMS-GT2 running on synthetic data for several PMS challenge instances demonstrate that it is competitive with both the EMS-GT and qPMS9, the two current best exact solutions for PMS. In particular, EMS-GT2 is able to reduce the run-times of EMS-GT by 18.33\%, 25.46\%, 19.88\% 14.81\% and 22.03\% for the (l,d) challenge instances (9, 2), (11, 3), (13, 4), (15, 5) and (17, 6) respectively. It also outperforms qPMS9, having runtime reductions of 94.41\%, 91.65\%, 85.40\%, 59.22\% and 5.8\% for the (9, 2), (11, 3), (13, 4), (15, 5) and (17, 6) synthetic challenge instances respectively.

	
\end{abstract}
